Peaks Over Threshold for Bursty Time Series

### 2018-04-24 article_springer/article_springer.Rmd

```{r get-rticles, eval=FALSE, include=FALSE, cache=TRUE}
# use this rticles version: 
devtools::install_github("strakaps/rticles", 
                         ref = "724aeb88d005e2f7befc37328633e7cb2bc1cce0")
# use the newest MittagLeffleR version:
devtools::install_github("strakaps/MittagLeffleR")
library(MittagLeffleR)
devtools::install_github("UNSW-MATH/CTRE")
library(CTRE)
```

@@

Time series displaying temporally inhomogeneous behaviour have received 
strong interest in the recent statistical physics literature 
[@Barabasi2005; @Oliveira2005; @Vasquez2006; @Vazquez2007; @Omi2011; 
@Min2010; @Karsai2011; @Bagrow2013]. They have been observed in the context 
of earthquakes, sunspots, neuronal activity and human communication [see 
@Karsai2012; @Vajna2013; @MeerschaertStoev08 for a list of references]. 
Such time series exhibit high activity in some 'bursty' intervals, which 
alternate with other, quiet intervals.  Although several mechanisms are 
plausible explanations for bursty behaviour (most prominently self-exciting
point processes [@hawkes1971point]), there seems to be one salient feature 
which very typically indicates the departure from temporal homogeneity: a 
heavy-tailed distribution of waiting times [@Vasquez2006; @Karsai2012; 
@Vajna2013]. As we show below in simulations, a simple renewal process with
heavy-tailed waiting times can capture this type of dynamics. For many 
systems, the renewal property is appropriate; a simple test of the absence 
of correlations in a succession of waiting times can be undertaken by 
randomly reshuffling the waiting times [@Karsai2012].

Often a magnitude, or mark can be assigned to each event in the renewal process, 
such as for earthquakes, solar flares or neuron voltages. 
Extreme value theory provides models for the distribution of the events
with the largest magnitude. 
A commonly made assumption in POT models is that times between 
events are either fixed 
or light-tailed, and this entails that the 
threshold crossing times form a Poisson process [@beirlantBook]. 
Then as one increases the threshold and thus decreases the threshold 
crossing probability $p$, the Poisson process is rarefied, i.e. its 
intensity decreases _linearly_ with $p$.

@@

We review the marked renewal process in Section 2, and 
derive a scaling limit theorem for inter-exceedance times in Section 3.
We give a statistical procedure to estimate model parameters via stability plots 
in Section 5, but to set the stage we first discuss inference for the 
Mittag-Leffler distribution in Section 4. 
Diagnostic plots for model criticism are discussed in Section 6. 
In Section 7, we discuss the non-memorylessness property of the 
Mittag-Leffler distribution, and how it affects the predictive distribution 
for the next threshold crossing event. 
Section 8 concludes.
For all statistical computations we have used R [@R]. 
Code used for the analysis in this article has been organized into 
an R package `CTRE` (<https://github.com/UNSW-MATH/CTRE>).
The source code for the figures generated in this manuscript is 
available online at <https://github.com/UNSW-MATH/bursty-POT>.

>>>>

```{r get-rticles, eval=FALSE, include=FALSE, cache=TRUE}
# use this rticles version: 
devtools::install_github("rstudio/rticles")
# use the newest MittagLeffleR version:
devtools::install_github("UNSW-MATH/MittagLeffleR")
library(MittagLeffleR)
devtools::install_github("UNSW-MATH/CTRE")
library(CTRE)
```

@@

Time series displaying temporally inhomogeneous behaviour have received 
strong interest in the recent statistical physics literature 
[@Barabasi2005; @Oliveira2005; @Vasquez2006; @Vazquez2007; @Omi2011; 
@Min2010; @Karsai2011; @Bagrow2013]. They have been observed in the context 
of earthquakes, sunspots, neuronal activity and human communication [see 
@Karsai2012; @Vajna2013; @MeerschaertStoev08 for a list of references]. 
Such time series exhibit high activity in some 'bursty' intervals, which 
alternate with other, quiet intervals.  Although several mechanisms are 
plausible explanations for bursty behaviour (most prominently self-exciting
point process by @hawkes1971point), there seems to be one salient feature 
which very typically indicates the departure from temporal homogeneity: a 
heavy-tailed distribution of waiting times [@Vasquez2006; @Karsai2012; 
@Vajna2013]. As we show below in simulations, a simple renewal process with
heavy-tailed waiting times can capture this type of dynamics. For many 
systems, the renewal property is appropriate; a simple test of the absence 
of correlations in a succession of waiting times can be undertaken by 
randomly reshuffling the waiting times [@Karsai2012].

Often a magnitude, or mark can be assigned to each event in the renewal process, 
such as for earthquakes, solar flares or neuron voltages. 
The Peaks-Over-Threshold model [POT, see e.g. @ColesBook] applies a
threshold to the magnitudes, and fits a Generalized Pareto distribution 
to the threshold exceedances. 
A commonly made assumption in POT models is that times between 
events are either fixed 
or light-tailed, and this entails that the 
threshold crossing times form a Poisson process [@Hsing88]. 
Then as one increases the threshold and thus decreases the threshold 
crossing probability $p$, the Poisson process is rarefied, i.e. its 
intensity decreases _linearly_ with $p$ [see e.g. @beirlantBook].

@@

We review the marked renewal process in Section 2, and 
derive a scaling limit theorem for inter-exceedance times in Section 3.
We give a statistical procedure to estimate model parameters via stability plots 
in Section 5, but to set the stage we first discuss inference for the 
Mittag-Leffler distribution in Section 4. 
(A simulation study of the effectiveness of our statistical procedure is 
given in the appendix.)
Diagnostic plots for model criticism are discussed in Section 6. 
In Section 7, we discuss the memory property of the 
Mittag-Leffler distribution, and how it affects the predictive distribution 
for the time until the next threshold crossing event. 
Section 8 concludes. 
For all statistical computations we have used R [@R]. 
All code and data used for the analysis in this article has been organized into 
an R package `CTRE` (<https://github.com/UNSW-MATH/CTRE>).
The source code for the figures generated in this manuscript is 
available online at <https://github.com/UNSW-MATH/bursty-POT>.
